Dispersion distance and the matter distribution of the Universe in dispersion space

### 2015-06-04 writeup/dm_space_distortions.tex
Edits to intro.

However, redshift is not the only measure of radial distance.   Standard
candles and standard rulers can be used to estimate distance using brightness
and angular size respectively.  In the same way a \emph{standard ping}, a brief
broadband radio impulse, can be used to estimate radial distance because of the
characteristic $\lambda^2$ dispersion of electromagnetic signals along the
line-of-sight as a function of wavelength.  This technique, routinely done with
pulsars in the Galaxy, can be used to define a \emph{dispersion distance} to
cosmological sources.  Intriguingly, the dispersion of fast radio bursts (FRB),
a new class of millisecond duration radio transients first reported by
\citet{2007Sci...318..777L}, has been measured with high significance.  These
bursts are the archetype of a standard ping with 10 reported to date
\citep{2012MNRAS.425L..71K,2013Sci...341...53T, 2014ApJ...790..101S,
2014ApJ...792...19B, 2015MNRAS.447..246P,2015ApJ...799L...5R}, and with a total
event rate estimated to be $3.3^{+5}_{-2.5} \times 10^3$ per sky per
day \cite{2015arXiv150500834R}.  Dispersion is characterized by a line-of-site
integral of the free electron density known the dispersion measure (DM).  FRB
events are observed to have dispersion measures of order $\sim\!\!1000 {\rm pc}
/ {\rm cm}^3$, greatly in excess of the galactic expectation of order
$\sim\!\!100\,{\rm pc} / {\rm cm}^3$ (depending on galactic latitude)
\cite{2014arXiv1412.4829D}.  They thus appear to be of extra-galactic origin,
with their dispersion produced either by the inter-galactic medium (IGM)
\citep{2013Sci...341...53T} surrounding galaxies along the line-of-sight
or the environment of the
source \citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.  If
the dispersion is dominated by electrons along the line-of-sight, then the
average dispersion--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I} and the
sources appear to be at cosmological distances of of order $\sim\!\!{\rm Gpc}$.
Such cosmological distances permit the study of clustering of FRB sources in
three dimensions with an expectation that these events trace the large-scale
structure of the Universe on linear scales. We note that while our discussion
here is framed in terms of FRBs, other bright radio transients that fluctuate
on timescales less than the typical differential lag of $\sim 1s/100 \,{\rm
MHz} $ might also be useful as standard pings.

@@

There is an advantage to the lack of kernel in the dispersion-space integral term.
Lensing tomography---the use of sources at multiple redshifts to un-mix
spatial scales---is complicated by the kernel. The kernel's shape depends on
the source redshift, re-weighting the line-of-sight integral and
making tomography inexact.  In contrast, the integrand in the dispersion-space integral
term is completely independent of the limit of integration, meaning
contributions from different parts of the line of sight (and thus different
spatial scales) can be separated exactly. A detailed study of dispersion space
tomography is beyond the scope of this work.  To avoid our results being
sensitive to the hard-to-model small scales, we assume that the contribution to the
integral term from radial distances $\chi'' < 500\,{\rm Mpc}/h$ are well
measured by the correlations with these radial slices. We thus neglect the
information from these slices in our sensitivity measurements and ignore
contributions to the integral term from below this distance. The smallest
spatial scale that contributes the our plotted angular power spectra out to
$\ell=1000$ is thus $k=2\,h/{\rm Mpc}$.

>>>>

However, redshift is not the only measure of radial distance. Standard
candles and standard rulers can be used to estimate distance using brightness
and angular size respectively.  In the same way a \emph{standard ping}, a brief
broadband radio impulse, can be used to measure dispersion through the
characteristic wavelength-squared delay of electromagnetic signals. The disperion is
proportional to the amount of plasma along the
line-of-sight and can thus be used to estimate radial distance.
This technique, routinely done with
pulsars in the Galaxy, can be used to define a \emph{dispersion distance} to
cosmological sources.
%
Dispersion is readily measured from fast radio bursts (FRB),
a new class of millisecond duration radio transients first reported by
\citet{2007Sci...318..777L}, which are the archetype of a standard ping.
Ten such bursts have been reported to date
\citep{2012MNRAS.425L..71K,2013Sci...341...53T, 2014ApJ...790..101S,
2014ApJ...792...19B, 2015MNRAS.447..246P,2015ApJ...799L...5R}, and the total
event rate has been estimated to be $3.3^{+5}_{-2.5} \times 10^3$ per sky per
day \cite{2015arXiv150500834R}.
Dispersion is characterized by a line-of-site
integral of the free electron density known the dispersion measure (DM).  FRB
events are observed to have dispersion measures of order $\sim\!\!1000\,{\rm pc}
/ {\rm cm}^3$, greatly in excess of the galactic expectation of order
$\sim\!\!100\,{\rm pc} / {\rm cm}^3$ (depending on galactic latitude)
\cite{2014arXiv1412.4829D}.  They thus appear to be of extra-galactic origin,
with their dispersion produced either by the inter-galactic medium (IGM)
surrounding galaxies along the line-of-sight \citep{2013Sci...341...53T}
or the environment of the
source \citep{2014ApJ...785L..26L,2015arXiv150101341P,2015arXiv150505535C}.  If
the dispersion is dominated by electrons along the line-of-sight, then the
average dispersion--distance relation can be modelled
\citep{2014ApJ...797...71Z,2004MNRAS.348..999I, 2003ApJ...598L..79I} and the
sources appear to be at cosmological distances of of order $\sim\!\!{\rm Gpc}$.
Such cosmological distances permit the study of clustering of FRB sources in
three dimensions with an expectation that these events trace the large-scale
structure of the Universe on linear scales. We note that while our discussion
here is framed in terms of FRBs, other bright radio transients that fluctuate
on timescales less than the typical differential lag of $\sim 1s/100 \,{\rm
MHz} $ might also be useful as standard pings.

@@

There is an advantage to the lack of kernel in the dispersion-space integral term.
Lensing tomography---the use of sources at multiple redshifts to un-mix
spatial scales---is complicated by the kernel. The kernel's shape depends on
the source redshift, re-weighting the line-of-sight integral and
making tomography inexact.  In contrast, the integrand in the dispersion-space integral
term is completely independent of the limit of integration, meaning
contributions from different parts of the line of sight (and thus different
spatial scales) can be separated exactly. A detailed study of dispersion space
tomography is beyond the scope of this work.  To avoid our results being
sensitive to the hard-to-model small scales, we assume that the contribution to the
integral term from radial distances $\chi'' < 500\,{\rm Mpc}/h$ are well
measured by the correlations with these radial slices. We thus ignore
contributions to the integral term from below this distance and neglect the
information from these slices in our sensitivity measurements. The smallest
spatial scale that contributes the our plotted angular power spectra out to
$\ell=1000$ is thus $k=2\,h/{\rm Mpc}$.
